This chapter provides an overview of existing systems and their relation to
Noria and partial state.

\section{Materialized Views}

Database materialized views~\cite{materialized-views} were originally devised to
store expensive analytical query results for quick recollection. Unfortunately,
commercial databases' materialized view support is limited, and views must
usually be rebuilt from scratch when the underlying data
changes~\cite{materialized-view-selection-sql-server,mssql-materialized-view-restrictions-blog,
mssql-materialized-view-restrictions}.

The key to usable materialized views is how they are maintained as the
underlying data changes. Rather than throw away the current materialized query
results, a good materialized view system should only perform the work needed to
\textit{incrementally} update the materialized results. This has been the
subject of considerable research in the past few decades.
\citeauthor{materialized-survey} gives a good survey of the current
landscape~\cite{materialized-survey}.

Modern incremental view maintenance (IVM) techniques tend to be based on
\textit{delta queries}, which are algebraically derived queries that give
efficient relational expressions for computing the change to a (materialized)
view given a set of changes to the underlying data. The current state-of-the-art
in IVM is \textit{Higher-Order IVM}, in which the system derives multiple,
recursive such delta queries for each view, and strategically materializes and
maintains the intermediate delta query results as well~\cite{dbtoaster, hotdog}.
Recent work also proposes techniques for mitigating the memory overhead of such
intermediate materializations by instead materializing smaller auxiliary state
from which the necessary values can then be efficiently produced when
needed~\cite{memory-efficient}. Sadly, few of these solutions have been adopted
in commercially available databases. Unlike Noria, these systems tend to focus
on long-term maintenance of analytics queries\,---\,they do not provide
mechanisms for fast reads and do not support eviction.

Noria's dataflow implements logic similar to Higher-Order IVM for each view,
including materialization of intermediate results. The algorithm Noria uses for
what dataflow to use to compute the changes to each view is comparatively naive,
and could likely benefit from the techniques in the aforementioned work.

Pequod~\cite{pequod} and DBProxy~\cite{dbproxy} provide materialized views that
also support partial materialization in response to client demand. However,
Pequod is limited to static queries specified in a datalog-like language, and
DBProxy does not support incremental view maintenance. And neither system shares
state nor processing across views.

\section{Caching}

Application-level caching is often implemented in an ad hoc fashion, and tends
to be the source of many application errors~\cite{ad-hoc-caching}. In
particular, such ad hoc system often fail to invalidate or update the cache as
the underlying data changes, leading to permanently stale entries. Researchers
and industry teams alike have therefore attempted to build systems to automate
cache maintenance.

Large-scale applications often end up building their own custom caching
infrastructure~\cite{facebook-memcache, flannel}, which solves their needs in
isolation, but does not provide a ready-to-use solution for other application
developers that face similar issues. These custom-built solutions also tend to
implement only the minimum functionality the authors need at the time, and
forego more complicated, but nonetheless useful features such as incremental
cache updates. Noria presents a ``plug and play'' solution specifically for
query result caching for many applications.

The research community has also produced several systems that aim to provide
more general-purpose transparent caching. TAO~\cite{tao} and
TxCache~\cite{txcache} implement automated query result caching, but do not
support incremental in-place cache updates like Noria.
CacheGenie~\cite{cachegenie} implements a trigger-based middleware cache for
object-relational mapping frameworks, and supports in-place cache updates, but
is limited to only specific operations supported by the framework. In contrast,
Noria transparently speeds up regular SQL queries, and does not require the
application to use a particular database abstraction framework.

In the world of database literature, database caching front ends are sometimes
referred to as \textit{Cache-Augmented SQL} systems. And there, like with all
caching systems, the primary concern is one of consistency\,---\,\emph{some}
mechanism must be in place to ensure that the cache remains up to date as the
underlying data changes. Research in this space tends to focus on augmenting the
key-value systems that stores cache entries so that the application can
correctly manage races between database updates and cache
invalidations~\cite{facebook-memcache, casql-consistency,
casql-consistency-thesis}. Noria instead integrates the cache management into
the database, which allows the cache entries to be incrementally updated,
automatically, in-place.

A related approach is \textit{mid-tier database caching}, in which subsets of
the database are replicated onto the hosts that run the application's code. This
allows certain queries to be run locally without interacting with the remote
database backend~\cite{mtcache}. While the approach is appealing in that some
database queries can avoid traversing the network, it does not provide the same
speedups that query result caching provides.

\section{Dataflow}

A wide range of dataflow and stream-processing systems exist that excel at
data-parallel computing~\cite{dryad, naiad, storm, heron, flink, millwheel,
spark-streaming, stanford-stream, s-store, cloud-dataflow}. However, these
systems cannot easily serve web applications directly. They only achieve
low-latency incremental updates at the expense of limiting how much state they
keep by \textit{windowing}, which results in incomplete results, or by keeping
full state in memory. Partial state allows Noria to lift this restriction.
Furthermore, these systems generally provide no mechanism for accessing computed
state except through the dataflow or by integrating with additional external
systems, which adds latency.

Many existing system are also limited to a fixed set of queries defined when the
system is started, and cannot easily adopt query changes as the application
evolves. Some dataflow systems do support Noria-like dynamic changes to the
running dataflow~\cite{ciel, ray}, but without support for demand-driven partial
state these systems must either fully compute results when the dataflow is
extended, or have new dataflow only take into account subsequent updates.

Some developers use, or consider using, a streaming fabric like Apache
Kafka~\cite{kafka} to build their own view maintenance pipeline~\cite{nyt-kafka,
samza-blogpost}. However, at the time of writing, no general-purpose system
exists based on such a pipeline that achieves the performance and flexibility of
Noria.

Differential dataflow~\cite{differential-dataflow}, and its instantiation in the
commercial product Materialize~\cite{materialize}, bears a striking resemblance
to Noria at first glance. In particular, it uses dataflow to produce
automatically-maintained materialized views over SQL queries. However,
Materialize does not implement partial state, and must therefore maintain
similar queries independently (which misses out on opportunities for shared
compute and state) or fully materialize query results (which uses more memory).
The authors behind Materialize have proposed partial solutions to some of these
challenges, which are discussed in \S\vref{s:disc:emulating}.
