\documentclass[fontsize=12pt,paper=letter]{scrartcl}

% for \sfrac
\usepackage{xfrac}

\begin{document}
\begin{titlepage}
  \begin{center}
    \textbf{\large Partial State in Dataflow-Based Materialized Views}\\
    \vspace{0.5\baselineskip}
    by\\
    \vspace{0.5\baselineskip}
    {\large Jon Gjengset}\\
    \vspace{0.5\baselineskip}
    \begin{addmargin}[0.15in]{0.15in}
      \centering
    Submitted to the Department of
    Electrical Engineering and Computer Science
    on \today{}
    in Partial Fulfillment of the Requirements for the Degree of
    Doctor of Philosophy in Computer Science
    \end{addmargin}
  \end{center}

  \begin{flushleft}
  ABSTRACT
  \vspace{0.5\baselineskip}

  This thesis proposes a practical database system that lowers latency and
    increases supported load for read-heavy applications by using
    incrementally-maintained materialized views to cache query results. As
    opposed to state-of-the-art materialized view systems, the presented system
    builds the cache on demand, and evicts cache entries in response to a
    shifting workload.

  \vspace{0.5\baselineskip}

  The enabling technique the thesis introduces is \textit{partially stateful
    materialization}, which allows entries in materialized views to be
    \textit{missing}. The thesis proposes \textit{upqueries} as a mechanism to
    fill such missing state on demand using dataflow, and implements them in the
    materialized view system Noria. The thesis then discusses additional
    mechanisms needed to establish eventual consistency for partially stateful
    dataflow.

  \vspace{0.5\baselineskip}

  Noria with partial materialization saves application developers from
    implementing their own ad hoc caching mechanisms to speed up their database
    accesses. Instead, the caching is built into the database, and is
    transparent to the application. Experimental results suggest that the
    presented system increases supported application load by up to $20\times$
    over MySQL and performs similarly to an optimized key-value store cache.
    Partial state also reduces memory use by up to $\sfrac{2}{3}$ compared to
    traditional materialized views.

  \vspace{\baselineskip}
  Thesis Supervisor: Robert Tappan Morris\\
  Title: Professor of Computer Science

  \vspace{0.5\baselineskip}
  Thesis Reader: M. Frans Kaashoek\\
  Title: Professor of Computer Science

  \vspace{0.5\baselineskip}
  Thesis Reader: Sam Madden\\
  Title: Professor of Computer Science

  \vspace{0.5\baselineskip}
  Thesis Reader: Malte Schwarzkopf\\
  Title: Assistant Professor, Brown University
  \end{flushleft}
\end{titlepage}
\end{document}
