This thesis is built on the belief that partial materialization is
useful. In this section, I provide experimental and qualitative evidence
that this belief is warranted through analysis of the following
questions:

\begin{enumerate}
 \item Why is materialization useful for real-world use-cases? \S\ref{s:eval:materialization}
 \item In what ways are existing materialization techniques lacking? \S\ref{s:eval:lacking}
 \item \textbf{How} does partial state improve the usefulness of materialization? \S\ref{s:eval:how}
 \item \textbf{When} does partial state improve the usefulness of materialization? \S\ref{s:eval:when}
 \item What is the cost of using partial state? \S\ref{s:eval:cost}
 \item How does partial state compare to existing solutions? \S\ref{s:eval:existing}
\end{enumerate}

I cover these questions in the following sections. The high-level
take-away is that partial state significantly reduces the labor cost and
memory overhead of application caching, is competitive with existing
non-partial solutions, and presents a trade-off between memory use and
increased tail latency.

\section{Why is materialization useful for real-world use-cases?}
\label{s:eval:materialization}

Caching
OSDI paper

\section{In what ways are existing materialization techniques lacking?}
\label{s:eval:lacking}

Caching is labor intensive + error prone (Facebook memcached paper)

DB view materialization (like DBToaster) ``all or nothing''

Commercial DB view materialization limited and slow (ref OSDI paper)

\section{How does partial state improve the usefulness of materialization?}
\label{s:eval:how}

By folding the cache into the database, database knows how to keep it up
to date — avoids having that logic threaded through the application.

Large parts of the application's dataset and queries are access
infrequently, if at all. We want to devote memory only where it matters.

Fig.: Lobsters memory use full/partial (many things not accessed) and
partial/evict (many things infrequently accessed).

Note that evict is necessary because, over time, more and more of the
tail will be sampled. Without eviction, eventually the entire tail would
be kept in memory (so == full).

Partial state also enables efficient migrations in many cases. Any view
that can be made partial (ref section + later eval) is immediately
available, and does not compute until requested.

Fig.: vote-migration skewed partial vs full

\section{When does partial state improve the usefulness of materialization?}
\label{s:eval:when}

\subsection{When does partial help?}

Materialization is generally useful when reads are more common than
writes, since it shifts the computation from reads to writes.
But, without partial state, materialization is an ``off/on'' option —
either you materialize all the results for a query, or none [TODO: not
quite true — more subtlety here — depends what you mean by ``query''
(e.g., parameterized or not)].

Partial state (and caching in general) enables \textbf{selective}
materialization. This makes sense if most application accesses are for
a particular subset of the dataset. That is, accesses are \textit{skewed}
towards particular data. In applications with skewed access patterns
(common — ref log-normal + Lobsters above), you can choose to only
materialize commonly-accessed results, rather than all. This uses
potentially far less memory (as indicated above), while still giving you
a speedup for \textbf{most} accesses. The more memory you provide the partial
system with, the more of your tail will be pre-computed, and the more of
your requests will be fast.

Fig.: mem limit vs latency [mean/50/95/99] in Lobsters.

Deciding what memory limit to set is challenging, in part because it
may change over time as the access patterns change, and in part because
it depends on load. The higher the load, the more requests will be in
the tail (and miss). Of course, more requests will be in the head too.

Fig.: throughput vs 95\%-ile latency [memlimits] in vote.

For example, for Zipf distribution, we can compute \%s:

Tab.: | load | skew | keys hit in 30s | keys hit by 99\% | 

This is (again) why eviction is important — without it, partial would approach
full.

Lobsters suggests that significant skew is common. Also ref other
studies on log-normal/zipf/skew.

\subsection{When can we use partial?}

\begin{itemize}
 \item SELECT COUNT(*) FROM table;
 \item Add ``karma'' query:
   \begin{itemize}
    \item If already have ``karma'' column, trivial in DB. But, equivalent in
      Noria to saying that query was there from the start, so no migration.
    \item If \textbf{not} already have ``karma'' column, DB has two choices:
     \begin{enumerate}
      \item Compute column == expensive, and same as Noria would do
      \item Make query compute value on-demand -- same cost, but repeated!
     \end{enumerate}
   \end{itemize}
 \item TopK
\end{itemize}

\section{What is the cost of using partial state?}
\label{s:eval:cost}

Migrations maybe? Latency timeline to show that early accesses are slow.
Maybe good place to have comparison with Redis? Maybe show vote uniform
numbers? Talk about ``warmup''.

Fig.: throughput vs latency w/wo partial in vote.

Note latency diff for lobsters, and that full fell over due to mem, and
no bigger machine w/o more cores available.

Note that the reason vote falls over when it does is single-core write
processing (since only one query, so one data-flow path).

\section{How does partial state compare to existing solutions?}
\label{s:eval:existing}

This is a difficult question to answer. High-performance solutions are
often developed specifically for a given application, and not available
as general-purpose tools (ref facebook memcache, need cites here).
And applying the general-purpose tools that \textbf{are} available (memcache,
redis) effectively, requires significant effort on the part of the
application authors (or the evaluators). To manually add caching support
to Lobsters' XX queries, including mitigation, thundering herd
mitigation, and incremental updates would be a massive undertaking.

In some sense, this alone is an argument for Noria's approach. Since the
database uses information it already possesses (in the form of
application queries) to automatically optimize accesses through
materialization, it is relatively easy to take advantage of the benefits
that Noria provides.

Nonetheless, to shed some light on the absolute performance that Noria
provides against a caching system, I include below a performance
comparison between Noria and Redis. To approximate how a carefully
planned and optimized application caching deployment might perform, it
runs a workload that is idealized for a caching system:

\begin{itemize}
 \item Every Redis access hits in cache, to emulate perfect thundering herd
   mitigation and invalidation-avoidance schemes.
 \item Nearly all accesses (99.9\%) are reads, since writes would be
   bottlenecked by the backing store.
 \item All accesses are for a single integer value, to emulate a system that has
   perfect cache coverage. Request access keys are chosen according to a Zipfian
    distribution with $\alpha = 1.08$ (moderate skew).
 \item Accesses are batched to reduce serialization cost and increase
   throughput. Specifically, reads are \texttt{MGET}s, and writes are pipelined
    \texttt{INCRBY}s.
\end{itemize}

This is not a realistic use of Redis, but it allows us to ``assume the best''
about the underlying caching strategy and system. The benchmark runs for four
minutes, and then samples latencies for another two minutes.

Fig.: throughput vs 95\%-ile latency [noria,redis]

Redis is single threaded, which necessarily limits its performance. If we assume
perfect sharding, Redis should be able to support 16 times the load on a 16-core
machine. We see that 16x Redis ~= Noria (??), which suggests that Noria is
competitive with manual caching schemes.
